---
title: "Assignment 1"
output: html_document
---

# variables 
POLY_ID : Census Block Group ID
MEDHVAL : Median value of all ownder occupied housing units
PCTBACHMOR : Proportion of residents in Block Group with at least a bachelor's degree
PCTVACANT : Proportion of housing units that are vacant
PCTSINGLES : Percent of housing units that are detached single family houses
NBELPOV100: Number of households with incomes below 100% poverty level (i.e., number of households living in poverty)
MEDHHINC : Median household income
```{r}
library(Hmisc)
library(DAAG)
library(car)  #to calculate VIF
library(MASS)
library(rsq)
library(tidyverse)
library(caret)
library(dplyr)
library(sf)
library(terra)
library(tmap)

install.packages(c("sf", "terra"))
install.packages("tmap")
```

```{r}
# load data and data frame

regression_data <- read.csv("RegressionData.csv")
attach(regression_data)
```

```{r}
# calculate summary statistics for MEDHVAL

hist(MEDHVAL)
mean(MEDHVAL)
sd(MEDHVAL)
```
```{r}
# calculate summary statistics for PCTBACHMOR

hist(PCTBACHMOR)
mean(PCTBACHMOR)
sd(PCTBACHMOR)
```

```{r}
# calculate summary statistics for NBELPOV100

hist(NBELPOV100)
mean(NBELPOV100)
sd(NBELPOV100)
```

```{r}
# calculate summary statistics for PCTVACANT

hist(PCTVACANT)
mean(PCTVACANT)
sd(PCTVACANT)
```

```{r}
# calculate summary statistics for PCTSINGLES

hist(PCTSINGLES)
mean(PCTSINGLES)
sd(PCTSINGLES)
```

```{r}
# create summary table

summary_table <- data.frame(
  Variable = c(
    "Dependent Variable",
    "MEDHVAL",
    "Predictors",
    "PCTBACHMOR",
    "MEDHHINC",
    "PCTVACANT",
    "PCTSINGLES",
    "NBELPOV100"
  ),
  Mean = c(
    NA,
    mean(MEDHVAL, na.rm = TRUE),
    NA,
    mean(PCTBACHMOR, na.rm = TRUE),
    mean(MEDHHINC, na.rm = TRUE),
    mean(PCTVACANT, na.rm = TRUE),
    mean(PCTSINGLES, na.rm = TRUE),
    mean(NBELPOV100, na.rm = TRUE)
  ),
  SD = c(
    NA,
    sd(MEDHVAL, na.rm = TRUE),
    NA,
    sd(PCTBACHMOR, na.rm = TRUE),
    sd(MEDHHINC, na.rm = TRUE),
    sd(PCTVACANT, na.rm = TRUE),
    sd(PCTSINGLES, na.rm = TRUE),
    sd(NBELPOV100, na.rm = TRUE)
  )
)

summary_table
```
```{r}
# check if any of the data have any zero values

sapply(regression_data[c("MEDHVAL", "PCTBACHMOR",
                         "PCTVACANT", "PCTSINGLES", "NBELPOV100")],
       function(x) sum(x == 0, na.rm = TRUE))
```

```{r}
# use the log command to create 5 new variables 

LNMEDHVAL <- log(MEDHVAL)
LNPCTBACHMOR <- log(1+PCTBACHMOR)
LNNBELPOV100 <- log(1+NBELPOV100)
LNPCTVACANT <- log(1+PCTVACANT)
LNPCTSINGLES <- log(1+PCTSINGLES)
```

```{r}
# view the histograms of the new variables 

hist(LNMEDHVAL)
hist(LNPCTBACHMOR)
hist(LNNBELPOV100)
hist(LNPCTVACANT)
hist(LNPCTSINGLES)
```
Note that the dependent variable does look more or less normal
after the transformation – hence, LNMEDHVAL will be used as
the dependent variable in the regression analysis. You will also
see that for the predictors, the logarithmic transformation only
helps normalize the NBELPOV100 variable (so we will use
LNNBELPOV100 in the subsequent analyses). The other
variables have a large spike at zero (i.e., zero-inflated
distributions) after the transformations, so we will use the
original, untransformed PCBACHMORE, PCTVACANT, and
PCTSINGLES variables in the regression

```{r}
# determine if the relationship between the dependent variables and each of the predictors is linear, create four scatter plots

plot(LNMEDHVAL,LNNBELPOV100)
plot(LNMEDHVAL,PCTBACHMOR)
plot(LNMEDHVAL,PCTVACANT)
plot(LNMEDHVAL,PCTSINGLES)
```

```{r}
# check the Pearson correlations between all of the predictors

cor(cbind(PCTVACANT, PCTSINGLES, PCTBACHMOR, LNNBELPOV100))
```
Note whether you observe severe multicollinearity, and whether it’s appropriate to include all 4 variables as predictors.

None of the variables show severe multicollinearity. The most severe result is -0.3197668. 

```{r}
# import the shape file

regression_data_shp <- st_read("RegressionDataFiles.shp")
```
```{r}
# create the map for LNMEDHVAL 

tmap_mode("plot")

map1 <- tm_shape(regression_data_shp) +
  tm_polygons(
    fill = "LNMEDHVAL",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 4,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Log Median Housing Value"
      ),
    lwd = 0.2
  ) +
  tm_title("Log Median Housing Value") +
  tm_layout(legend.outside = TRUE)

map1
```
```{r}
# create the predictor maps

map2 <- tm_shape(regression_data_shp) +
  tm_polygons(
    fill = "PCTVACANT",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 4, values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Vacant Housing"
      ),
    lwd = 0.1
  ) +
  tm_title("Percent Vacant")

map3 <- tm_shape(regression_data_shp) +
  tm_polygons(
    fill = "PCTSINGLES",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 4, values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Single-Family Housing"
      ),
    lwd = 0.1
  ) +
  tm_title("Percent Single-Family")

map4 <- tm_shape(regression_data_shp) +
  tm_polygons(
    fill = "PCTBACHMOR",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 4, values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Bachelor's Degree or Higher"
      ),
    lwd = 0.1
  ) +
  tm_title("Percent Bachelor's Degree or Higher")

regression_data_shp$LNNBELPOV100 <- LNNBELPOV100
map5 <- tm_shape(regression_data_shp) +
  tm_polygons(
    fill = "LNNBELPOV100",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 4, values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Log Households Below Poverty"
      ),
    lwd = 0.1
  ) +
  tm_title("Log Households Below Poverty")

map2
map3
map4
map5
```
```{r}
# create figures for the predictor maps

predictor_maps <- tmap_arrange(
  map2, map3, map4, map5,
  ncol = 2,
  nrow = 2
)

predictor_maps

# save the figures

tmap_save(
  predictor_maps,
  filename = "predictor_maps.png",
  width = 10,
  height = 8,
  units = "in",
  dpi = 300
)

# save LNMEDHVAL figure

tmap_save(
  map1,
  filename = "LNMEDHVAL_map.png",
  width = 8,
  height = 6,
  units = "in",
  dpi = 300
)

```

```{r}
# regression analysis 

regression <- lm(LNMEDHVAL ~ PCTVACANT + PCTSINGLES + PCTBACHMOR + LNNBELPOV100, data = regression_data)

regression
```

```{r}
# summary table

summary(regression)
```

```{r}
# ANOVA table

anova(regression)
```

```{r}
regression_predicted <- fitted(regression)
regression_residuals <- residuals(regression)
regression_standardresiduals <- rstandard(regression)

regression_predicted
regression_residuals
regression_standardresiduals
```

```{r}
# scatter plot of standardized residuals 

plot(
  regression_predicted,
  regression_standardresiduals,
  xlab = "Predicted Values",
  ylab = "Standardized Residuals",
  main = "Standardized Residuals vs. Predicted Values",
  pch = 20
)

abline(h = 0, lty = 2, col = "red")
```

```{r}
# stepwise regression

step_regression <- stepAIC(regression, direction = "both")
step_regression
step_regression$anova
AIC(step_regression)
```

The stepwise regression procedure using the Akaike Information Criterion (AIC) selected the initial model as the final model. All four predictors—PCTVACANT, PCTSINGLES, PCTBACHMOR, and LNNBELPOV100—were retained because removing any individual predictor increased the AIC. The full model had an AIC of -3448.16, the lowest among the models evaluated. Therefore, the procedure did not identify an improvement by removing any of the predictors.

```{r}
# k fold validation 
regression_data$LNNBELPOV100 <- LNNBELPOV100
cv1 <- CVlm(
  data = regression_data,
  form.lm = formula(regression),
  m = 5,
  plotit = FALSE
)

# format
# Create plot data
plot_data <- data.frame(
  Observed = cv1$LNMEDHVAL,
  Predicted = cv1$cvpred,
  Fold = factor(cv1$fold)
)

# Assign a different color to each fold
fold_colors <- c("steelblue", "tomato", "forestgreen", "orange", "purple")

# Leave space on the right for the legend
par(mar = c(5, 5, 4, 10), xpd = TRUE)

# Plot observed versus predicted values
plot(
  plot_data$Observed,
  plot_data$Predicted,
  pch = 16,
  cex = 0.6,
  col = fold_colors[as.numeric(plot_data$Fold)],
  xlab = "Observed log median housing value",
  ylab = "Cross-validated predicted value",
  main = "5-Fold Cross-Validation",
  bty = "l"
)

# Reference line for perfect predictions
abline(0, 1, lty = 2, lwd = 2, col = "gray40")

# Legend on the right
legend(
  "topright",
  inset = c(-0.65, 0),
  legend = c(paste("Fold", 1:5), "Perfect prediction"),
  col = c(fold_colors, "gray40"),
  pch = c(rep(16, 5), NA),
  lty = c(rep(NA, 5), 2),
  lwd = c(rep(NA, 5), 2),
  bty = "n",
  cex = 0.8
)

par(xpd = FALSE)


rmse1 <- sqrt(mean(
  (cv1$LNMEDHVAL - cv1$cvpred)^2,
  na.rm = TRUE
))

rmse1
```

```{r}
# second cross validation 

# Create the second regression model with only two predictors
regression2 <- lm(
  LNMEDHVAL ~ PCTVACANT + MEDHHINC,
  data = regression_data
)

# Run 5-fold cross-validation
cv2 <- CVlm(
  data = regression_data,
  form.lm = formula(regression2),
  m = 5,
  plotit = FALSE
)

# Format the cross-validation results
plot_data <- data.frame(
  Observed = cv2$LNMEDHVAL,
  Predicted = cv2$cvpred,
  Fold = factor(cv2$fold)
)

# Assign a different color to each fold
fold_colors <- c("steelblue", "tomato", "forestgreen", "orange", "purple")

# Leave space on the right for the legend
par(mar = c(5, 5, 4, 10), xpd = TRUE)

# Plot observed versus predicted values
plot(
  plot_data$Observed,
  plot_data$Predicted,
  pch = 16,
  cex = 0.6,
  col = fold_colors[as.numeric(plot_data$Fold)],
  xlab = "Observed log median housing value",
  ylab = "Cross-validated predicted value",
  main = "5-Fold Cross-Validation: PCTVACANT + MEDHHINC",
  bty = "l"
)

# Reference line for perfect predictions
abline(0, 1, lty = 2, lwd = 2, col = "gray40")

# Legend on the right
legend(
  "topright",
  inset = c(-0.65, 0),
  legend = c(paste("Fold", 1:5), "Perfect prediction"),
  col = c(fold_colors, "gray40"),
  pch = c(rep(16, 5), NA),
  lty = c(rep(NA, 5), 2),
  lwd = c(rep(NA, 5), 2),
  bty = "n",
  cex = 0.8
)

par(xpd = FALSE)

# Calculate RMSE
rmse2 <- sqrt(mean(
  (cv2$LNMEDHVAL - cv2$cvpred)^2,
  na.rm = TRUE
))

rmse2
```

cross validation 1 RMSE: 0.3664401
cross validation 2 RMSE: 0.4427216

```{r}
hist(regression_standardresiduals)
```

```{r}
regression_data_shp$regression_standardresiduals <-
  rstandard(regression)
map6 <- tm_shape(regression_data_shp) +
  tm_polygons(
    fill = "regression_standardresiduals",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 4, values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Standardized Regression Residuals"
      ),
    lwd = 0.1
  ) +
  tm_title("Standardized Regression Residuals")

map6
```
## REPORT - Draft by Wednesday, Submit Friday 

## INTRODUCTION (Sara)

## METHODS (Gautami)
# Data Cleaning
# Exploratory Data Analysis
# Multiple Regression Analysis 
# Additional Analysis 
# Software

## RESULTS (Tegan)
# Exploratory Results
# Regression Results
# Regression Assumption Checks
# Additional Models

## DISCUSSION & LIMITATIONS (Sara)



