---
title: "Assignment 1"
output: html_document
---

# variables 
POLY_ID : Census Block Group ID
MEDHVAL : Median value of all ownder occupied housing units
PCTBACHMOR : Proportion of residents in Block Group with at least a bachelor's degree
PCTVACANT : Proportion of housing units that are vacant
PCTSINGLES : Percent of housing units that are detached single family houses
NBELPOV100: Number of households with incomes below 100% poverty level (i.e., number of households living in poverty)
MEDHHINC : Median household income

```{r}
# load data and data frame

regression_data <- read.csv("RegressionData.csv")
attach(regression_data)
```

```{r}
# calculate summary statistics for MEDHVAL

hist(MEDHVAL)
mean(MEDHVAL)
sd(MEDHVAL)
```
```{r}
# calculate summary statistics for PCTBACHMOR

hist(PCTBACHMOR)
mean(PCTBACHMOR)
sd(PCTBACHMOR)
```

```{r}
# calculate summary statistics for NBELPOV100

hist(NBELPOV100)
mean(NBELPOV100)
sd(NBELPOV100)
```

```{r}
# calculate summary statistics for PCTVACANT

hist(PCTVACANT)
mean(PCTVACANT)
sd(PCTVACANT)
```

```{r}
# calculate summary statistics for PCTSINGLES

hist(PCTSINGLES)
mean(PCTSINGLES)
sd(PCTSINGLES)
```

```{r}
# create summary table

summary_table <- data.frame(
  Variable = c(
    "Dependent Variable",
    "MEDHVAL",
    "Predictors",
    "PCTBACHMOR",
    "MEDHHINC",
    "PCTVACANT",
    "PCTSINGLES",
    "NBELPOV100"
  ),
  Mean = c(
    NA,
    mean(MEDHVAL, na.rm = TRUE),
    NA,
    mean(PCTBACHMOR, na.rm = TRUE),
    mean(MEDHHINC, na.rm = TRUE),
    mean(PCTVACANT, na.rm = TRUE),
    mean(PCTSINGLES, na.rm = TRUE),
    mean(NBELPOV100, na.rm = TRUE)
  ),
  SD = c(
    NA,
    sd(MEDHVAL, na.rm = TRUE),
    NA,
    sd(PCTBACHMOR, na.rm = TRUE),
    sd(MEDHHINC, na.rm = TRUE),
    sd(PCTVACANT, na.rm = TRUE),
    sd(PCTSINGLES, na.rm = TRUE),
    sd(NBELPOV100, na.rm = TRUE)
  )
)

summary_table
```
```{r}
# check if any of the data have any zero values

sapply(regression_data[c("MEDHVAL", "PCTBACHMOR",
                         "PCTVACANT", "PCTSINGLES", "NBELPOV100")],
       function(x) sum(x == 0, na.rm = TRUE))
```

```{r}
# use the log command to create 5 new variables 

LNMEDHVAL <- log(MEDHVAL)
LNPCTBACHMOR <- log(1+PCTBACHMOR)
LNNBELPOV100 <- log(1+NBELPOV100)
LNPCTVACANT <- log(1+PCTVACANT)
LNPCTSINGLES <- log(1+PCTSINGLES)
```

```{r}
# view the histograms of the new variables 

hist(LNMEDHVAL)
hist(LNPCTBACHMOR)
hist(LNNBELPOV100)
hist(LNPCTVACANT)
hist(LNPCTSINGLES)
```
Note that the dependent variable does look more or less normal
after the transformation – hence, LNMEDHVAL will be used as
the dependent variable in the regression analysis. You will also
see that for the predictors, the logarithmic transformation only
helps normalize the NBELPOV100 variable (so we will use
LNNBELPOV100 in the subsequent analyses). The other
variables have a large spike at zero (i.e., zero-inflated
distributions) after the transformations, so we will use the
original, untransformed PCBACHMORE, PCTVACANT, and
PCTSINGLES variables in the regression

```{r}
# determine if the relationship between the dependent variables and each of the predictors is linear, create four scatter plots

plot(LNMEDHVAL,LNNBELPOV100)
plot(LNMEDHVAL,PCTBACHMOR)
plot(LNMEDHVAL,PCTVACANT)
plot(LNMEDHVAL,PCTSINGLES)
```

```{r}
# check the Pearson correlations between all of the predictors

cor(cbind(PCTVACANT, PCTSINGLES, PCTBACHMOR, LNNBELPOV100))
```
Note whether you observe severe multicollinearity, and whether it’s appropriate to include all 4 variables as predictors.

None of the variables show severe multicollinearity. The most severe result is -0.3197668 which is not severe?


